E. coli resistance patterns from Community Resistance in Athens Project

### 2021-10-29 products/manuscript/Manuscript.Rmd
Part 3: Modeling

## Exploratory analysis

#load needed packages. make sure they are installed.
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving

library(dplyr) #for data wrangling
library(reshape) #for reshaping and melting data

@@

#Let's look at demographics for the study.
mydata %>% ggplot(aes(x = race)) +
  geom_bar()
#^The race data shows that a majority of the participants are white.
mydata %>% ggplot(aes(x = biological_sex)) +
  geom_bar()
#^This biological sex data shows that, when observing males and females,
#the participant group is primarily female.
mydata %>% ggplot(aes(x = numeric_age)) +
  geom_bar()
#^This age data shows that the age of the participants skews
#strongly left/young.
```

@@

#Let's look at the food data. I want to make a separate data set of
#food information so I can make a melted data set.
#The figure at the end shows the number of participants that ate
#a typ

@@

abs_pattern <- mydata %>% subset(select=c(28:33)) %>%
  mutate(amp_proportion_resistant = amp_resistant_isolates / total_isolates) %>%
  mutate(ceft_proportion_resistant = ceft_resistant_isolates / total_isolates) %>%
  mutate(cipro_proportion_resistant = cipro_resistant_isolates / total_isolates) %>%
  mutate(tetra_proportion_resistant = tetra_resistant_isolates / total_isolates) %>%
  mutate(trimet_proportion_resistant = trimet_resistant_isolates / total_isolates)

@@

#Wow, there kind of seem to be patterns here, but also in other ways there
#is no correlation. I had a thought that we could remove any zero values, but 
#that is a really bad idea because it is straight up removing data to make 
#a pattern, so I will not do that. You can kind of see a diagonal line, but
#I do not think we can say anything because of the excessive outliers.

#I want to see if ampicillin resistance and number of days an individual ate
#fruit that week are correlated

@@

@@

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example table \@ref(tab:resulttable) shows a table summarizing a linear model fit.

```{r resulttable,  echo=FALSE}
resulttable=readRDS("../../results/resulttable.rds")
knitr::kable(resulttable, caption = 'Linear model fit table.')
```

>>>>

## Exploratory analysis
```{r}
###############################
# analysis script
#
#this script loads the processed, cleaned data, does a simple analysis
#and saves the results to the results folder

#load needed packages. make sure they are installed.
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(dplyr) #for data wrangling
library(reshape) #for reshaping and melting data
library(tidymodels) #for modeling

@@

#Let's look at demographics for the study.
mydata %>% ggplot(aes(x = race)) +
  geom_bar()
#^The race data shows that a majority of the participants are white.
mydata %>% ggplot(aes(x = biological_sex)) +
  geom_bar()
#^This biological sex data shows that, when observing males and females,
#the participant group is primarily female.
mydata %>% ggplot(aes(x = numeric_age)) +
  geom_bar()
#^This age data shows that the age of the participants skews
#strongly left/young.

#histogram for samples with greater than or equal to one isolate showing resistance 
#to a given antibiotic

mydata[c("amp_resistant_isolates")] %>% filter(amp_resistant_isolates >= 1) %>% 
  ggplot(aes(x=amp_resistant_isolates)) + geom_histogram() #ampicillin

mydata[c("ceft_resistant_isolates")] %>% filter(ceft_resistant_isolates >= 1) %>% 
  ggplot(aes(x=ceft_resistant_isolates)) + geom_histogram() #ceftriaxone

mydata[c("cipro_resistant_isolates")] %>% filter(cipro_resistant_isolates >= 1) %>% 
  ggplot(aes(x=cipro_resistant_isolates)) + geom_histogram() #ciprofloxacin

mydata[c("tetra_resistant_isolates")] %>% filter(tetra_resistant_isolates >= 1) %>% 
  ggplot(aes(x=tetra_resistant_isolates)) + geom_histogram() #tetracycline

mydata[c("trimet_resistant_isolates")] %>% filter(trimet_resistant_isolates >= 1) %>% 
  ggplot(aes(x=trimet_resistant_isolates)) + geom_histogram() #trimethoprim

#There is not a strong pattern amongst any of these.On an intitial glance,
#ciprofloxacin and ceftriaxone are skewed left and the rest are skewed right.

#Examine seasonality of resistance to different antibiotics,
#starting with ampicillin
mydata %>% ggplot(aes(y=amp_resistant_isolates, x = date_produced)) +  geom_point()
#ceftriaxone
mydata %>% ggplot(aes(y=ceft_resistant_isolates, x = date_produced)) +  geom_point()
#ciprofloxacin
mydata %>% ggplot(aes(y=cipro_resistant_isolates, x = date_produced)) +  geom_point()
#tetracycline
mydata %>% ggplot(aes(y=tetra_resistant_isolates, x = date_produced)) +  geom_point()
#trimethoprim
mydata %>% ggplot(aes(y=trimet_resistant_isolates, x = date_produced)) +  geom_point()
```

@@

#Let's look at the food data. I want to make a separate data set of
#food information so I can make a melted data set.

@@

abs_pattern <- mydata %>%
  mutate(amp_proportion_resistant = amp_resistant_isolates / total_isolates) %>%
  mutate(ceft_proportion_resistant = ceft_resistant_isolates / total_isolates) %>%
  mutate(cipro_proportion_resistant = cipro_resistant_isolates / total_isolates) %>%
  mutate(tetra_proportion_resistant = tetra_resistant_isolates / total_isolates) %>%
  mutate(trimet_proportion_resistant = trimet_resistant_isolates / total_isolates)
abs_pattern
```

@@

#Wow, there kind of seem to be patterns here, but also in other ways thereis no correlation. I had a thought that we could remove any zero values, but that is a really bad idea because it is straight up removing data to make a pattern, so I will not do that. You can kind of see a diagonal line, but I do not think we can say anything because of the excessive outliers.

#I want to see if ampicillin resistance and number of days an individual ate fruit that week are correlated

@@

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._
```{r}

@@

lm_mod <- linear_reg() %>% set_engine("lm")

#Fits another linear model to the continuous outcome
#(ampicillin resistance proportion) using all (important) predictors of interest.
ampprop_and_predictors <- abs_pattern %>%
  subset(select = c(amp_proportion_resistant,
                    regular_healthcare_exposure,
                    healthcare_exposure_yn,
                    regular_animal_exposure,
                    has_pets,
                    cat_andor_dog,
                    untreated_recreational_water_exposure_past_week,
                    untreated_recreational_water_exposure_past_month,
                    treated_recreational_water_exposure_past_month,
                    treated_recreational_water_exposure_past_week,
                    eat_poultry_past_week,
                    eat_pork_or_beef_past_week,
                    eat_fish_or_shellfish_past_week,
                    eat_dairy_past_week,
                    eat_raw_fruits_or_vegetables_past_week))
ampprop_and_predictors

lm_ampall_model <- lm_mod %>% fit(amp_proportion_resistant ~ ., data = ampprop_and_predictors)
ampallpredictors_lm <- tidy(lm_ampall_model)
ampallpredictors_lm

ampallpredictors_table <- broom::tidy(lm_ampall_model)
ampallpredictors_table

#Looking at this, only dairy seems to have statistical significance. 
#Again, these are p-values, so approach with some skepticism.

abs_pattern %>% ggplot(aes(x = eat_dairy_past_week, y = amp_proportion_resistant)) +
  geom_point()
amp_dairy_yn_plot <- abs_pattern %>% mutate(any_dairy_week_yn = eat_dairy_past_week > 0) %>% 
  ggplot(aes(x = any_dairy_week_yn, y = amp_proportion_resistant)) +
  geom_boxplot()

#plotting but only for isolates that show SOME ampicillin resistance.
#I'm just doing this for exploration's sake.
abs_pattern %>% mutate(any_dairy_week_yn = eat_dairy_past_week > 0) %>% 
  filter(amp_proportion_resistant > 0) %>%
  ggplot(aes(x = any_dairy_week_yn, y = amp_proportion_resistant)) +
  geom_boxplot() #the mean is a lot higher

#Tetracycline
tetprop_and_predictors <- abs_pattern %>%
  subset(select = c(tetra_proportion_resistant,
                    regular_healthcare_exposure,
                    healthcare_exposure_yn,
                    regular_animal_exposure,
                    has_pets,
                    cat_andor_dog,
                    untreated_recreational_water_exposure_past_week,
                    untreated_recreational_water_exposure_past_month,
                    treated_recreational_water_exposure_past_month,
                    treated_recreational_water_exposure_past_week,
                    eat_poultry_past_week,
                    eat_pork_or_beef_past_week,
                    eat_fish_or_shellfish_past_week,
                    eat_dairy_past_week,
                    eat_raw_fruits_or_vegetables_past_week))
tetprop_and_predictors

lm_tetall_model <- lm_mod %>% fit(tetra_proportion_resistant ~ ., data = tetprop_and_predictors)
tetallpredictors_lm <- tidy(lm_tetall_model)
tetallpredictors_lm
#no significance for tetracycline

tetallpredictors_table <- broom::tidy(lm_tetall_model)
tetallpredictors_table

#Trimethoprim
triprop_and_predictors <- abs_pattern %>%
  subset(select = c(trimet_proportion_resistant,
                    regular_healthcare_exposure,
                    healthcare_exposure_yn,
                    regular_animal_exposure,
                    has_pets,
                    cat_andor_dog,
                    untreated_recreational_water_exposure_past_week,
                    untreated_recreational_water_exposure_past_month,
                    treated_recreational_water_exposure_past_month,
                    treated_recreational_water_exposure_past_week,
                    eat_poultry_past_week,
                    eat_pork_or_beef_past_week,
                    eat_fish_or_shellfish_past_week,
                    eat_dairy_past_week,
                    eat_raw_fruits_or_vegetables_past_week))
triprop_and_predictors

lm_triall_model <- lm_mod %>% fit(trimet_proportion_resistant ~ ., data = triprop_and_predictors)
triallpredictors_lm <- tidy(lm_triall_model)
triallpredictors_lm
#no significance for trimethoprim

triallpredictors_table <- broom::tidy(lm_triall_model)
triallpredictors_table

#Fits another linear model to the 
#categorical outcome using all (important) predictors of interest.
log_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

abs_pattern_yndairy <- abs_pattern %>% mutate(any_dairy_week_yn = eat_dairy_past_week > 0)

#linear regression between ampicillin and dairy (as y/n)
ampdairy_lm_fit <-lm_mod %>%
  fit(amp_proportion_resistant ~ any_dairy_week_yn, data = abs_pattern_yndairy)
ampdairy_lm <- tidy(ampdairy_lm_fit)
ampdairy_lm

ampallpredictors_table
tetallpredictors_table
triallpredictors_table

ampallpredictors_figure = here("results", "ampallpredictors_table")
saveRDS(ampallpredictors_table, file = ampallpredictors_figure)

tetallpredictors_figure = here("results", "tetallpredictors_table")
saveRDS(tetallpredictors_table, file = tetallpredictors_figure)

triallpredictors_figure = here("results", "triallpredictors_table")
saveRDS(triallpredictors_table, file = triallpredictors_figure)
```
